---
title: "Class 12 Lab Session"
author: 'Nicole Alfonso, (PID: A16429176)'
date: "2024-02-15"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Proportion of G/G in a population 

Here we are reading the CSV file 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

Looking at the type of genotypes. 
```{r}
table(mxl$Genotype..forward.strand.)
```

Calculating the frequency of each genotype in the Mexican Ancestry in Los Angeles sample population (MXL)
```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

# Section 4. Population Analysis 

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

First, I will import the data

```{r}
snp <- "https://bioboot.github.io/bimm143_W24/class-material/rs8067378_ENSG00000172057.6.txt"
expr <- (read.table(snp))
head(expr)
```

Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

```{r}
nrow(expr)
table(expr$geno)
summary(expr)
```
According to the output of the summary function, the median output of their expression levels is 30.14

Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch = TRUE)
```

