---
title: "Class05: Data Visualization w ggplot"
author: "Nicole Alfonso (A16429176)"
format: pdf
---

# Graphics systems in R 

There are many graphics systems in R for making plots and figures.

We have already played a little with **"base R"** graphics and the `plot()` function.  

Today we will start learning about a popular graphics package called `ggplot2()`. 

This is an add on package - i.e. we need to install it. I install it like I install any package, with the `install.packages()` function. 

```{r}
plot(cars)
```

Before I can use the functions from a package I have to load up the package from my "library". We use the `library(ggplot2)` command to load it up. You also have to add extra stuff to ggplot. 

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is made up of at least 3 things: 
- data (the numbers etc. that will go into your plot)
- aesthetics (how the columns of the data map to the plot aesthetics 
geoms (how the plot actually looks, points, bars, lines, etc. )

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist)+ 
  geom_point() 
```

For simple plots, ggplot is more verbose - it takes more code than base R plot. 

Add some more layers to our plot 
```{r}
ggplot(cars) + 
  aes(x=speed, y=dist)+ 
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Stopping distnace of old cars", 
       subtitle = "A silly example plot") +
theme_minimal()
```

```{r}

#File location online 

url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)

nrow(genes)
colnames(genes)
ncol(genes)
table(genes$State)
(127/5196)*100

```


```{r}
p <- ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) + 
  geom_point() + 
  labs(title = "Gene Expression Changes Upon Drug Treatment", x="Control (no drug)", y="Drug Treatment") 
p + scale_color_manual(values=c("blue", "gray", "red")) 
```

Optional Extension

Question 7 

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"
gapminder <- read.delim(url)
install.packages("gapminder")
library(gapminder)
# install.packages("dplyr")
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)
```

```{r}
ggplot(gapminder_2007) + 
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) + 
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007) + 
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) + 
  geom_point(alpha=0.8)
```

```{r}
ggplot(gapminder_2007) + 
  aes(x=gdpPercap, y=lifeExp, color=pop) + 
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10)
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year==2007)
ggplot(gapminder_1957) + 
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) + 
  geom_point(alpha=0.7) + 
  scale_size_area(max_size = 10) + 
  facet_wrap(~year)
```

Question 8: Bar Chart 

```{r}
gapminder_top5 <- gapminder %>% filter (year==2007) %>% arrange(desc(pop)) %>% top_n(5, pop)

gapminder_top5
```

```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x=country, y=pop))
```

```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x=country, y=pop, fill=continent))
```

```{r}
ggplot(gapminder_top5) + 
  geom_col(aes(x=country, y=pop, fill=lifeExp))
```

```{r}
ggplot(gapminder_top5) + 
  (aes(x=reorder(country, -pop), y=pop, fill=country)) + 
  geom_col(col="gray30") + 
  guides(fill="none")
```